Positivity-Preserving Adaptive Runge--Kutta Methods

### 2020-04-30 nusslein_vsrp/latex/overview.tex
Revise remark on Example II step size.

We take the linear, positivity preserving ODE
\cite{kopecz_unconditionally_2018}
\begin{equation}
  u'(t) = L u(t),
  \quad
  u(0) = \begin{pmatrix} 1 \\ 0 \end{pmatrix},
  \qquad
  L = \begin{pmatrix} -5 & 1 \\ 5 & -1\end{pmatrix},
\end{equation}
and use the three stage, third order SSP method SSP(3,3)
\begin{align}
\begin{array}{c|ccc}
0 &  &  & \\
1 & 1 &  & \\
\nicefrac{1}{2} & \nicefrac{1}{4} & \nicefrac{1}{4} & \\
\hline
 & \nicefrac{1}{6} & \nicefrac{1}{6} & \nicefrac{2}{3}\\
\end{array}
\end{align}
of \cite{shu1988efficient}.
The matrix $L$ has the eigenvalues zero and $-6$ and its operator
norm is $2 \sqrt{13}$.
The real stability interval of SSP33 includes $[-2.5,0]$. Hence,
it seems reasonable\footnote{A more careful analysis following
\cite{ranocha2018L2stability} shows that the method is not
monotone (energy stable) for this problem and choice of $\dt$ since
$L$ is not normal.} to choose $\dt = \nicefrac{1}{3}$.
The corresponding stage derivatives are
\begin{equation}
  f(y_1) = \begin{pmatrix} -5 \\ 5 \end{pmatrix},\quad
  f(y_2) = \begin{pmatrix} 5 \\ -5 \end{pmatrix},\quad
  f(y_3) = \begin{pmatrix} -5 \\ 5 \end{pmatrix}.
\end{equation}
The values of the next step using the standard weights is
\begin{equation}
  u^1 = \begin{pmatrix} \nicefrac{-1}{9} \\ \nicefrac{10}{9} \end{pmatrix}.
\end{equation}
Since the first component of the new solution is negative,
we want to adapt the weights to ensure positivity.
All weights that comply with the constraints for first and second
order of accuracy can be expressed as
\begin{equation}
  \bt =
  \begin{pmatrix}
    \nicefrac{1}{6} \\
    \nicefrac{1}{6} \\
    \nicefrac{2}{3}
  \end{pmatrix}
  + \alpha \begin{pmatrix}
    \nicefrac{1}{2} \\
    \nicefrac{1}{2} \\
    -1
  \end{pmatrix},
  \qquad
  \alpha \in \R.
\end{equation}
We have one degree of freedom for the choice of the weights,
parameterized by $\alpha$.
If the general expression for the weights is inserted in
\eqref{eq:Combination} the general solution is
\begin{equation}
  u^{1}
  =
  u^0 + \dt \left(f_1, f_2, f_3\right) \bt
  =
  \begin{pmatrix}
    \nicefrac{-1}{9} \\
    \nicefrac{10}{9}
  \end{pmatrix}
  +\alpha \begin{pmatrix}
    5 \\
    -5
  \end{pmatrix}.
\end{equation}
By changing the parameter $\alpha$, the weights and the new
solution are altered. With a suitable choice of
$\alpha \in \left[ \nicefrac{1}{45}, \nicefrac{2}{9} \right]$,
any $u$ that complies with mass conservation and positivity can
be reached.
By adding additional constraints on the weights, the choice of
$\alpha$ can be narrowed down.
An objective function is also needed to make the choice unique. This should be designed in a way to prefer weights that are close to the original weights.

>>>>

We take the linear, positivity preserving ODE
\cite{kopecz_unconditionally_2018}
\begin{equation}
  u'(t) = L u(t),
  \quad
  u(0) = \begin{pmatrix} 1 \\ 0 \end{pmatrix},
  \qquad
  L = \begin{pmatrix} -5 & 1 \\ 5 & -1\end{pmatrix},
\end{equation}
and use the three stage, third order SSP method SSP(3,3)
\begin{align}
\begin{array}{c|ccc}
0 &  &  & \\
1 & 1 &  & \\
\nicefrac{1}{2} & \nicefrac{1}{4} & \nicefrac{1}{4} & \\
\hline
 & \nicefrac{1}{6} & \nicefrac{1}{6} & \nicefrac{2}{3}\\
\end{array}
\end{align}
of \cite{shu1988efficient}.
The matrix $L$ has the eigenvalues zero and $-6$ and its operator
norm is $2 \sqrt{13}$.
The real-axis stability interval of SSP33 includes the interval $[-2.5,0]$.
We take $\dt = \nicefrac{1}{3}$, which satisfies the spectral
condition and guarantees boundedness (though not monotonicity)
of the solution.
The corresponding stage derivatives are
\begin{equation}
  f(y_1) = \begin{pmatrix} -5 \\ 5 \end{pmatrix},\quad
  f(y_2) = \begin{pmatrix} 5 \\ -5 \end{pmatrix},\quad
  f(y_3) = \begin{pmatrix} -5 \\ 5 \end{pmatrix}.
\end{equation}
The values of the next step using the standard weights is
\begin{equation}
  u^1 = \begin{pmatrix} \nicefrac{-1}{9} \\ \nicefrac{10}{9} \end{pmatrix}.
\end{equation}
Since the first component of the new solution is negative,
we want to adapt the weights to ensure positivity.
All weights that comply with the constraints for first and second
order of accuracy can be expressed as
\begin{equation}
  \bt =
  \begin{pmatrix}
    \nicefrac{1}{6} \\
    \nicefrac{1}{6} \\
    \nicefrac{2}{3}
  \end{pmatrix}
  + \alpha \begin{pmatrix}
    \nicefrac{1}{2} \\
    \nicefrac{1}{2} \\
    -1
  \end{pmatrix},
  \qquad
  \alpha \in \R.
\end{equation}
We have one degree of freedom for the choice of the weights,
parameterized by $\alpha$.
If the general expression for the weights is inserted in
\eqref{eq:Combination} the general solution is
\begin{equation}
  u^{1}
  =
  u^0 + \dt \left(f_1, f_2, f_3\right) \bt
  =
  \begin{pmatrix}
    \nicefrac{-1}{9} \\
    \nicefrac{10}{9}
  \end{pmatrix}
  +\alpha \begin{pmatrix}
    5 \\
    -5
  \end{pmatrix}.
\end{equation}
By changing the parameter $\alpha$, the weights and the new
solution are altered. With a suitable choice of
$\alpha \in \left[ \nicefrac{1}{45}, \nicefrac{2}{9} \right]$,
any $u$ that complies with mass conservation and positivity can
be reached.
By adding additional constraints on the weights, the choice of
$\alpha$ can be narrowed down.
An objective function is also needed to make the choice unique. This should be designed in a way to prefer weights that are close to the original weights.
